Improved Hierarchical Clustering Through Nesting

### 2015-05-30 icdm/submission.tex
Typos

\begin{figure*}[!t]
\centerline{\subfloat[Partitional Clustering]{\includegraphics[width =1.75in]{partitional_clustering}
\label{fig_part}}
\hfil
\subfloat[Hierarchical Clustering]{\includegraphics[width=2.5in]{hier_clustering}
\label{fig_second_case}}}
\caption{A partitional clustering for $k=3$ and a hierarchical clustering of the data where the clustering on the third level down (represented by the red line) corresponds to the partitional clustering on the left.}
\label{fig_hier}
\end{figure*}

@@

Given solution pairs $(S_1, a_1)$ and $(S_2, a_2)$, we say that $(S_1, a_1)$ is \emph{nested} in $(S_2, a_2)$ (denoted $(S_1, a_1) \subseteq (S_2, a_2)$) if the following conditions hold:
\begin{enumerate}[\IEEEsetlabelwidth{3)}]
	\item $S_1 \subset S_2$, 
	\item $\forall  j \in X$, if $a_2(j) \in S_1$, then $a_1(j) = a_2(j)$, 
	\item $\forall j, k \in X$, if $a_2(j) = a_2(k)$, then $a_1(j) = a_1(k)$.
\end{enumerate}
A solution to the hierarchical $k$-median problem is a nested chain of solutions, and the hierarchical clustering it generates is found by looking at the clusters given by each $(S_i, a_i)$. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the hierarchy. An $\alpha$-approximation algorithm for this problem generates a nested chain of solutions in polynomial time such that for all $k = 1, \ldots, |X|$ there exists $(S_i, a_i)$ in the chain such that $|S_i| \leq k$ and $f(S_i, a_i) \leq \alpha \cdot f(O)$ for all $O \subseteq X$ with $|O| = k$.

@@

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. In particular, their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to greedily construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. Using the local $p$-swap $(3+\frac{2}{p})$-approximation algorithm mentioned above as the black-box $k$-median solver performed well in experimental hierarchical $k$-median test problems \cite{Nagarajan}. While the authors tested the variants of this algorithm for different $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

Several algorithms have been built upon agglomerative clustering that we choose not to compare with our proposed algorithms. Divisive clustering, similar to agglomerative, starts with the data in one cluster and repeatedly splits a cluster on each level. This method has not been studied nearly as much as agglomerative clustering \cite{ESL} but has had relative success in classifying document datasets \cite{Zhao}.  Algorithms such as the CURE, ROCK, and Chameleon \cite{Guha_CURE, Guha_ROCK, Karypis} incorporate non-spherical cluster shapes and cluster density, but do not relate as much to the above approximation algorithm in their objective. We also choose not compare the approximation framework with Bayesian hierarchical methods since these methods require many more parameters and add additional prior information that our algorithms do not use.

@@

From each non-empty bucket, pick the solution with the smallest number of cluster centers, and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r$. We then inductively build our nested solutions. Set $S_1 = \overline{V}_1$ and $a_1$ to be the assignment that assigns each point to the closest point in $S_1$. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we construct $S_i$ by  adding the closest point $j \in S_{i-1}$ to $l$ for each $l \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

@@

\subsection{Partitional Clustering Algorithms}
This nesting framework allows us to choose the partitional clustering algorithm $\mathcal{A}$. As mentioned above, PAM and $k$-medoids are two potential algorithms. PAM has the added benefit of being a 5-approximation algorithm for the partitional $k$-median problem which yields to an overall approximation ratio of $20.71 \cdot 5 = 103.55$ for the hierarchical $k$-median problem ($10.03 \cdot 5 = 50.15$ randomized). We also considered the size of the local swaps in PAM to improve the approximation ratio but found that the approximation ratio of constructed solutions for given values of $k$ did not significantly improve while the computation time increased greatly.

We also consider two linear programming based algorithms, one of which has an approximation guarantee. We first present the linear programming formulation for the $k$-median problem. Suppose we have a set $X$ and we let $s_i \in \{0,1\}$ be the variable representing whether or not we want to make $i$ a center for all $i \in X$. We also create a variable $a_{i,j} \in \{0, 1\}$ which represents whether or not we assign $i \in X$ to the cluster with center $j \in X$. Given this set and assignment, the $k$-median objective function is 
\begin{equation}
\sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j), 
\end{equation}
which we want to minimize. However, we are restricted by the fact that we can choose at most $k$ centers, every point must be assigned to a single cluster center, and that we can only assign a point to a center that is actually in $S$. This is represented by the three equations:
\begin{eqnarray}
\sum_{i \in X} s_i   \leq k & ,\\
\sum_{j \in X} a_{i,j}  = 1 & \forall i \in X, \\
a_{i,j}   \leq s_j & \forall i, j \in X.
\end{eqnarray}
Therefore, solving the following integer program would directly find the optimal set of centers and assignment for the $k$-median problem.
\begin{eqnarray}
\text{Min } & \sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j) \\
\text{s.t. } & \sum_{i \in X} s_i   \leq k  \\
& \sum_{j \in X} a_{i,j}  = 1 \quad \forall i \in X \\
& a_{i,j}   \leq s_j \quad \forall i, j \in X \\
& s_i \in \{0,1\} \quad \forall i \in X \\
& a_{i,j} \in \{0,1\} \quad \forall i, j \in X
\end{eqnarray}
Relaxing the constraints so that $s_i \in [0,1]$ for all $i \in X$ and $a_{i,j} \in [0,1]$ for all $i,j \in X$ yields the linear programming (LP) relaxation for the $k$-median problem, which we can solve in polynomial time.

Interestingly, the solution to the linear programming is often integer valued \cite{Nagarajan}. Based on this fact, we propose two partitional clustering algorithms: one that solves the LP relaxation and constructs $S$ by taking the points in $X$ with the $k$ highest $s_i$ values and one that solves the LP relaxation and constructs $S$ by choosing $i$ to be in $S$ with probability $s_i$, sampling for each $i \in X$ until $k$ centers are chosen. We refer to these algorithms as LP-greedy and LP-rand, respectively.
Charikar and Li \cite{Charikar} introduce a 3.25-approximation algorithm that uses this LP solution to smartly construct a set of centers. We refer to this algorithm as LP-Charikar. We choose to implement this LP algorithm rather than the LP-based approximation algorithm with best approximation ratio because the algorithm is simpler to implement (and we expect it to have a lower running time).

@@

All tests were conducted using Python 2.7.9 on a 64-bit Windows platform with a 4 core 3.4 Ghz Intel i7 processor and 16 GB physical memory. All code is available at \url{https://github.com/apaul29/Hierarchical-Code}.
We chose to test our methods on a variety of data:
\begin{LaTeXdescription}
\item[Gaussian] 27 synthetic $k$-center Gaussian data sets, generated in Python. In particular, Gauss\textunderscore d\textunderscore k\textunderscore e was generated by first choosing $k$ random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \cdot \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] \cite{Beasley} From a collection of classic test data sets for approximation algorithms created by Beasley with 40 test sets of size $100, 200, \ldots, 900$ and varying values of $k$. This data is unlabeled.
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{LaTeXdescription}

\begin{table}[!t]
\caption{Some specifics about our datasets}
\label{table_specifics}
\centering
\begin{tabular}{ | l | l | l | l | l | }
\hline
 Dataset Name & Opt Obj & $k$ & $n$ & dim \\ \hline
Soybean & 114 & 4 & 47 & 35 \\ 
Iris & 98 & 3 & 150 & 4 \\ 
Gauss\_5\_5\_5 & 195.4 & 5 & 388 & 5 \\ 
Gauss\_5\_10\_5 & 270 & 10 & 544 & 5 \\ 
Gaussian avg & 241 & 22 & 507 & 5 \\ 
pmed5 & 1355 & 33 & 100 & \  \\ 
pmed20 & 1789 & 133 & 400 & \  \\ 
% pmed40 & 5128 & 90 & 900 & \  \\ 
$p$-median avg & 5535 & 46 & 460 & \  \\ \hline
 \end{tabular}
\end{table}

@@

Before implementing these data sets in the hiearchical framework, we tested them as partitional algorithms on the data above which had given values of $k$. For comparison, we also included the results from running $k$-means (implemented in the scikit-learn package in Python) and the extension of PAM called 2-swaps that also looks at local swaps of size 2 (which improves the $k$-median approximation ratio to 4 from 5).

For each data set and each partitional clustering method, we recorded the approximation ratio (for the flat $k$-median objective function), computation time, and, where applicable, adjusted mutual information (AMI).  AMI is a measure which scores accuracy of predicted clusters compared to a true labeling by measuring the agreement of two assignments. The measure is normalized to reflect chance, so that a perfect labeling receives a score of 1 and an independent random labeling receives a score of 0 in expectation.  See \cite{Vinh} for further information.  Results are given in Tables~\ref{fig:flat_hier_comp}-\ref{fig:flat_hier_approx}  below with some results averaged due to space constraints. The results reported are average values across 10 trials.

\begin{table*}[!t]
\label{fig:flat_results_comp}
\caption{Average Computation Time (sec) for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l|}
\hline
Dataset	&	k-means	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	0.0151	&	0.0020	&	0.0159	&	0.0493	&	0.1180	&		&		\\
Iris	&	0.0129	&	0.0030	&	0.1018	&	0.2294	&	1.4644	&		&		\\
Gauss\_5\_5\_5	&	0.0249	&	0.0318	&	1.5058	&	2.3824	&	18.1701	&		&		\\
Gauss\_5\_10\_5	&	0.0540	&	0.1085	&	5.4071	&	9.4277	&	47.7316	&		&		\\
Gauss\_5\_50\_5	&	0.1371	&	0.3019	&	21.4457	&	39.6144	&	17.5115	&		&		\\
Gaussian avg	&	0.0730	&	0.1817	&	13.0214	&	18.6957	&	37.2315	&		&		\\
pmed5	&	N/A	&	0.0381	&	0.6223	&	0.7479	&	0.5357	&		&		\\
pmed20	&	N/A	&	1.7947	&	44.9791	&	68.2306	&	18.2055	&		&		\\
pmed35	&	N/A	&	0.1133	&	5.8814	&	9.2881	&	117.4038	&		&		\\
$p$-median avg	&	N/A	&	0.8072	&	40.3474	&	56.7595	&	46.1541	&		&		\\ \hline
\end{tabular}
\end{table*}

\begin{table*}[!t]
\label{fig:flat_results_AMI}
\caption{Average AMI measure for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l| }
\hline
Dataset	&	k-means	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	0.686	&	0.550	&	0.732	&	0.708	&	0.721	&		&		\\
Iris	&	0.748	&	0.690	&	0.760	&	0.757	&	0.647	&		&		\\
Gauss\_5\_5\_5	&	0.741	&	0.687	&	0.749	&	0.744	&	0.754	&		&		\\
Gauss\_5\_10\_5	&	0.783	&	0.677	&	0.725	&	0.737	&	0.729	&		&		\\
Gauss\_5\_50\_5	&	0.646	&	0.621	&	0.619	&	0.631	&	0.631	&		&		\\
Gaussian avg	&	0.777	&	9.112	&	0.761	&	0.761	&	0.767	&		&		\\ \hline
\end{tabular}
\end{table*}

\begin{table*}[!t]
\label{fig:flat_results_approx}
\caption{Average Approximation Coeffiecients for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l| l|}
\hline
Dataset	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	1.167	&	1.012	&	1.012	&	1.000	&		&		\\
Iris	&	1.081	&	1.004	&	1.004	&	0.873	&		&		\\
Gauss\_5\_5\_5	&	1.070	&	1.002	&	1.005	&	1.002	&		&		\\
Gauss\_5\_10\_5	&	1.080	&	1.011	&	1.009	&	1.020	&		&		\\
Gauss\_5\_50\_5	&	1.089	&	1.020	&	1.014	&	1.001	&		&		\\
Gaussian avg	&	1.064	&	1.009	&	1.008	&	1.004	&		&		\\
pmed5	&	1.373	&	1.026	&	1.033	&	1.000	&		&		\\
pmed20	&	1.501	&	1.032	&	1.027	&	1.000	&		&		\\
pmed35	&	1.083	&	1.005	&	1.007	&	1.027	&		&		\\
$p$-median avg	&	1.187	&	1.017	&	1.019	&	1.000	&		&		\\ \hline
\end{tabular}
\end{table*}

@@

Results are given in Tables~\ref{fig:hier_results_comp}-\ref{fig:hier_results_approx} for the full hierarchical version and in Tables~\ref{fig:hier_log_results_compt}-\ref{fig:hier_log_results_approx} for the logarithmic version. The results reported are average values across 10 trials.

\begin{table*}[!t]
\label{fig:hier_results_comp}
\caption{Average Computation Time (sec) for Hierarchical Clustering Algorithms. Besides the agglomerative methods, the algorithm is denoted by the $k$-median solver used and [R] and [D] are used to denote the randomized and determinisitic versions, respectively.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l| l| }
\hline
Dataset	&	Ward Linkage	&	Average Linkage	&	$k$-medoids [R]	&	$k$-medoids [D]	&	LP-Charikar [R]	&	LP-rand [R]	&	LP-greedy [R]	&	LP-greedy [D]	\\ \hline
Soybean	&	0.0002	&	0.0002	&	0.2132	&	0.2086	&	0.9903	&	0.4068	&	0.2817	&	0.4708	\\
Iris	&	0.0011	&	0.0009	&	3.2308	&	3.4530	&	32.3171	&	11.5128	&	8.8181	&	14.0168	\\
Gauss\_5\_5\_5	&	0.0123	&	0.0105	&	165.3442	&	106.6187	&	712.6960	&	273.1223	&	267.1039	&	222.1711	\\
Gauss\_5\_10\_5	&	0.0284	&	0.0260	&	556.8065	&	383.8541	&	2088.8757	&	793.7941	&	622.8916	&	623.8670	\\
Gauss\_5\_50\_5	&	0.0121	&	0.0112	&	176.8477	&	128.2697	&	764.3297	&	269.5327	&	253.2762	&	222.7860	\\
Gaussian avg	&	0.0250	&	0.0217	&	521.3331	&	354.4588	&	1695.7987	&	666.5034	&	566.4133	&	567.0086	\\
pmed5	&	N/A	&	0.0004	&	1.4797	&	1.1839	&		&	3.8239	&	4.6917	&		\\
pmed20	&	N/A	&	0.0098	&	153.5358	&	121.9013	&	550.8604	&	240.8364	&	193.8426	&	189.6854	\\
pmed35	&	N/A	&	0.0666	&	3225.4469	&	1957.6519	&		&	1939.0170	&	2326.1185	&	1801.2031	\\
$p$-median avg	&	N/A	&	0.0243	&	1021.8706	&	652.1928	&		&	723.1452	&	857.3750	&		\\ \hline
\end{tabular}
\end{table*}

>>>>

\begin{figure*}[!]
\centerline{\subfloat[Partitional Clustering]{\includegraphics[width =1.75in]{partitional_clustering}
\label{fig_part}}
\hfil
\subfloat[Hierarchical Clustering]{\includegraphics[width=2.5in]{hier_clustering}
\label{fig_hier}}}
\caption{A partitional clustering for $k=3$, where the colors represent the different partitions, and a dendrogram representation of a hierarchical clustering of the data where edges represent merges of clusters and the clustering on the third level down (represented by the red line) corresponds to the partitional clustering on the left.}
\end{figure*}

@@

Given solution pairs $(S_1, a_1)$ and $(S_2, a_2)$, we say that $(S_1, a_1)$ is \emph{nested} in $(S_2, a_2)$ (denoted $(S_1, a_1) \subseteq (S_2, a_2)$) if the following conditions hold:
\begin{enumerate}[\IEEEsetlabelwidth{3)}]
	\item $S_1 \subset S_2$, 
	\item $\forall  j \in X$, if $a_2(j) \in S_1$, then $a_1(j) = a_2(j)$, 
	\item $\forall j, k \in X$, if $a_2(j) = a_2(k)$, then $a_1(j) = a_1(k)$.
\end{enumerate}
A solution to the hierarchical $k$-median problem is a nested chain of solutions, and the hierarchical clustering it generates is found by looking at the clusters given by each $(S_i, a_i)$. An example is given in Fig.~\ref{fig_chain1}-\ref{fig_chain4}. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the hierarchy. An $\alpha$-approximation algorithm for this problem generates a nested chain of solutions in polynomial time such that for all $k = 1, \ldots, |X|$ there exists $(S_i, a_i)$ in the chain such that $|S_i| \leq k$ and $f(S_i, a_i) \leq \alpha \cdot f(O)$ for all $O \subseteq X$ with $|O| = k$.

\begin{figure*}[b]
\centering
\subfloat[$(S_1, a_1)$]{\includegraphics[width =1.75in]{chain1}
\label{fig_chain1}} \hspace{15mm}
\subfloat[$(S_2, a_2)$]{\includegraphics[width=1.75in]{chain2}
\label{fig_chain2}} \\
\subfloat[$(S_3, a_3)$]{\includegraphics[width=1.75in]{chain3}
\label{fig_chain3}}  \hspace{15mm}
\subfloat[$(S_4, a_4)$]{\includegraphics[width=1.75in]{chain4}
\label{fig_chain4}}
\caption{An example of a nested chain of solutions. $S_i$ is indicated by the colored nodes and $a_i$ by the edges. The clusters on that level are given by the connected components. Note that $S_{i-1}$ is a subset of $S_i$ and the assignments stay consistent. }
\label{fig_sim}
\end{figure*}

@@

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. In particular, their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to greedily construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. Using the local $p$-swap $(3+\frac{2}{p})$-approximation algorithm mentioned above as the black-box $k$-median solver performed well in experiments on hierarchical $k$-median test problems \cite{Nagarajan}. While the authors tested the variants of this algorithm for different $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

Several algorithms have been built upon agglomerative clustering that we choose not to compare with our proposed algorithms. Divisive clustering, similar to agglomerative, starts with the data in one cluster and repeatedly splits a cluster on each level. This method has not been studied nearly as much as agglomerative clustering \cite{ESL} but has had relative success in classifying document datasets \cite{Zhao}.  Algorithms such as the CURE, ROCK, and Chameleon \cite{Guha_CURE, Guha_ROCK, Karypis} incorporate non-spherical cluster shapes and cluster density, but do not relate as much to the above approximation algorithm in their objective. We also choose not to compare our algorithms with Bayesian hierarchical methods since these methods require many more parameters and add additional prior information that our algorithms do not use.

@@

From each non-empty bucket, pick the solution with the smallest number of cluster centers, and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r$. We then inductively build our nested solutions: Set $S_1 = \overline{V}_1$ and $a_1$ to be the assignment that assigns each point to the closest point in $S_1$. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we construct $S_i$ by  adding the closest point $j \in S_{i-1}$ to $l$ for each $l \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

@@

\subsection{Partitional Clustering Algorithms}
This nesting framework allows us to choose the partitional clustering algorithm $\mathcal{A}$. As mentioned above, PAM and $k$-medoids are two potential algorithms. PAM has the added benefit of being a 5-approximation algorithm for the partitional $k$-median problem which yields to an overall approximation ratio of $20.71 \cdot 5 = 103.55$ for the hierarchical $k$-median problem ($10.03 \cdot 5 = 50.15$ randomized). We looked at increasing the size of the local swaps in PAM to improve the approximation ratio but found that the approximation ratio of constructed solutions for given values of $k$ did not significantly improve while the computation time increased greatly.

We also consider three linear programming based algorithms, one of which has an approximation guarantee. We first present the linear programming formulation for the $k$-median problem. Suppose we have a set $X$. Let $s_i \in \{0,1\}$ be the variable representing whether or not we want to make $i$ a center for all $i \in X$. We also create a variable $a_{i,j} \in \{0, 1\}$ which represents whether or not we assign $i \in X$ to the cluster with center $j \in X$. Given this set and assignment, the $k$-median objective function is 
\begin{equation}
\sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j), 
\end{equation}
which we want to minimize. However, we are restricted by the fact that we can choose at most $k$ centers, every point must be assigned to a single cluster center, and that we can only assign a point to a center that is actually in $S$. This is represented by the following three equations, respectively:
\begin{eqnarray}
\sum_{i \in X} s_i   \leq k & ,\\
\sum_{j \in X} a_{i,j}  = 1 & \forall i \in X, \\
a_{i,j}   \leq s_j & \forall i, j \in X.
\end{eqnarray}
Therefore, solving the following integer program would directly find the optimal set of centers and assignment for the $k$-median problem.
\begin{eqnarray}
\text{Min } & \sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j) \\
\text{s.t. } & \sum_{i \in X} s_i   \leq k  \\
& \sum_{j \in X} a_{i,j}  = 1 \quad \forall i \in X \\
& a_{i,j}   \leq s_j \quad \forall i, j \in X \\
& s_i \in \{0,1\} \quad \forall i \in X \\
& a_{i,j} \in \{0,1\} \quad \forall i, j \in X
\end{eqnarray}
Relaxing the constraints so that $s_i \in [0,1]$ for all $i \in X$ and $a_{i,j} \in [0,1]$ for all $i,j \in X$ yields the linear programming (LP) relaxation for the $k$-median problem, which we can solve in polynomial time.

Interestingly, the solution to the linear programming is often integer valued \cite{Nagarajan}. Based on this fact, we propose two LP-based clustering algorithms: one that solves the LP relaxation and constructs $S$ by taking the points in $X$ with the $k$ highest $s_i$ values and one that solves the LP relaxation and constructs $S$ by choosing $i$ to be in $S$ with probability $s_i$, sampling for each $i \in X$ until $k$ centers are chosen. We refer to these algorithms as LP-greedy and LP-rand, respectively.
Charikar and Li \cite{Charikar} introduce a 3.25-approximation algorithm that uses this LP solution to smartly construct a set of centers. We refer to this algorithm as LP-Charikar. We choose to implement this LP algorithm rather than the LP-based approximation algorithm with best approximation ratio because the algorithm is simpler to implement (and we expect it to have a lower running time).

@@

All tests were conducted using Python 2.7.9 on a 64-bit Windows platform with a 4 core 3.4 Ghz Intel i7 processor and 16 GB physical memory. All code and the full computational results are available at \url{https://github.com/apaul29/Hierarchical-Code}.
We chose to test our methods on data from a few different contexts:
\begin{LaTeXdescription}
\item[Gaussian] 27 synthetic $k$-center Gaussian data sets, generated in Python. In particular, Gauss\textunderscore d\textunderscore k\textunderscore e was generated by first choosing $k$ random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \cdot \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] \cite{Beasley} From a collection of classic test data sets for approximation algorithms created by Beasley with 40 test sets of size $100, 200, \ldots, 900$ and varying values of $k$. This data is unlabeled.
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{LaTeXdescription}

\begin{table}[!t]
\caption{Some specifics about our datasets}
\label{table_specifics}
\centering
\begin{tabular}{ | l | l | l | l | l | }
\hline
 Dataset Name & Opt Obj & $k$ & $n$ & dim \\ \hline
Soybean & 113.805 & 4 & 47 & 35 \\ 
Iris & 98.214 & 3 & 150 & 4 \\ 
Gauss\_5\_5\_5 & 195.414 & 5 & 388 & 5 \\ 
Gauss\_5\_10\_5 & 270.066 & 10 & 544 & 5 \\ 
Gauss\_5\_50\_5 & 178.122 & 50 & 408 & 5 \\ 
Gaussian avg & 241.497 & 21.7 & 507 & 5 \\ 
pmed5 & 1355 & 33 & 100 & \  \\ 
pmed20 & 1789 & 133 & 400 & \  \\ 
pmed35 & 10400 & 5 & 800 & \  \\ 
$p$-median avg & 5535.3 & 45.9 & 460 & \  \\ \hline
 \end{tabular}
\end{table}

@@

Before implementing these data sets in the hierarchical framework, we compared them as partitional algorithms on the data above which had given values of $k$. For context, we also included the results from running $k$-means (implemented in the scikit-learn package in Python) and the extension of PAM called 2-swaps that also looks at local swaps of size 2 (which improves the $k$-median approximation ratio to 4 from 5).

For each data set and each partitional clustering method, we recorded the approximation ratio (for the flat $k$-median objective function), computation time, and, where applicable, adjusted mutual information (AMI).  AMI is a measure which scores accuracy of predicted clusters compared to a true labeling by measuring the agreement of two assignments. The measure is normalized to reflect chance, so that a perfect labeling receives a score of 1 and an independent random labeling receives a score of 0 in expectation.  See \cite{Vinh} for further information.  Results are given in Tables~\ref{fig:flat_results_comp}-\ref{fig:flat_results_approx}  below with some results averaged due to space constraints. The results reported are average values across 10 trials.

\begin{table*}[!t]
\caption{Average Computation Time (sec) for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l|}
\hline
Dataset	&	k-means	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	0.0151	&	0.0020	&	0.0159	&	0.0493	&	0.1180	&		&		\\
Iris	&	0.0129	&	0.0030	&	0.1018	&	0.2294	&	1.4644	&		&		\\
Gauss\_5\_5\_5	&	0.0249	&	0.0318	&	1.5058	&	2.3824	&	18.1701	&		&		\\
Gauss\_5\_10\_5	&	0.0540	&	0.1085	&	5.4071	&	9.4277	&	47.7316	&		&		\\
Gauss\_5\_50\_5	&	0.1371	&	0.3019	&	21.4457	&	39.6144	&	17.5115	&		&		\\
Gaussian avg	&	0.0730	&	0.1817	&	13.0214	&	18.6957	&	37.2315	&		&		\\
pmed5	&	N/A	&	0.0381	&	0.6223	&	0.7479	&	0.5357	&		&		\\
pmed20	&	N/A	&	1.7947	&	44.9791	&	68.2306	&	18.2055	&		&		\\
pmed35	&	N/A	&	0.1133	&	5.8814	&	9.2881	&	117.4038	&		&		\\
$p$-median avg	&	N/A	&	0.8072	&	40.3474	&	56.7595	&	46.1541	&		&		\\ \hline
\end{tabular}
\label{fig:flat_results_comp}
\end{table*}

\begin{table*}[!t]
\caption{Average AMI measure for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l| }
\hline
Dataset	&	k-means	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	0.686	&	0.550	&	0.732	&	0.708	&	0.721	&		&		\\
Iris	&	0.748	&	0.690	&	0.760	&	0.757	&	0.647	&		&		\\
Gauss\_5\_5\_5	&	0.741	&	0.687	&	0.749	&	0.744	&	0.754	&		&		\\
Gauss\_5\_10\_5	&	0.783	&	0.677	&	0.725	&	0.737	&	0.729	&		&		\\
Gauss\_5\_50\_5	&	0.646	&	0.621	&	0.619	&	0.631	&	0.631	&		&		\\
Gaussian avg	&	0.777	&	9.112	&	0.761	&	0.761	&	0.767	&		&		\\ \hline
\end{tabular}
\label{fig:flat_results_AMI}
\end{table*}

\begin{table*}[!t]
\caption{Average Approximation Coefficients for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l| l|}
\hline
Dataset	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	1.167	&	1.012	&	1.012	&	1.000	&		&		\\
Iris	&	1.081	&	1.004	&	1.004	&	0.873	&		&		\\
Gauss\_5\_5\_5	&	1.070	&	1.002	&	1.005	&	1.002	&		&		\\
Gauss\_5\_10\_5	&	1.080	&	1.011	&	1.009	&	1.020	&		&		\\
Gauss\_5\_50\_5	&	1.089	&	1.020	&	1.014	&	1.001	&		&		\\
Gaussian avg	&	1.064	&	1.009	&	1.008	&	1.004	&		&		\\
pmed5	&	1.373	&	1.026	&	1.033	&	1.000	&		&		\\
pmed20	&	1.501	&	1.032	&	1.027	&	1.000	&		&		\\
pmed35	&	1.083	&	1.005	&	1.007	&	1.027	&		&		\\
$p$-median avg	&	1.187	&	1.017	&	1.019	&	1.000	&		&		\\ \hline
\end{tabular}
\label{fig:flat_results_approx}
\end{table*}

@@

Results are given in Tables~\ref{fig:hier_results_comp}-\ref{fig:hier_results_approx} for the full hierarchical version and in Tables~\ref{fig:hier_log_results_comp}-\ref{fig:hier_log_results_approx} for the logarithmic version. The results reported are average values across 10 trials.

\begin{table*}[!t]
\caption{Average Computation Time (sec) for Hierarchical Clustering Algorithms. Besides the agglomerative methods, the algorithm is denoted by the $k$-median solver used and [R] and [D] are used to denote the randomized and deterministic versions, respectively.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l| l| }
\hline
Dataset	&	Ward Linkage	&	Average Linkage	&	$k$-medoids [R]	&	$k$-medoids [D]	&	LP-Charikar [R]	&	LP-rand [R]	&	LP-greedy [R]	&	LP-greedy [D]	\\ \hline
Soybean	&	0.0002	&	0.0002	&	0.2132	&	0.2086	&	0.9903	&	0.4068	&	0.2817	&	0.4708	\\
Iris	&	0.0011	&	0.0009	&	3.2308	&	3.4530	&	32.3171	&	11.5128	&	8.8181	&	14.0168	\\
Gauss\_5\_5\_5	&	0.0123	&	0.0105	&	165.3442	&	106.6187	&	712.6960	&	273.1223	&	267.1039	&	222.1711	\\
Gauss\_5\_10\_5	&	0.0284	&	0.0260	&	556.8065	&	383.8541	&	2088.8757	&	793.7941	&	622.8916	&	623.8670	\\
Gauss\_5\_50\_5	&	0.0121	&	0.0112	&	176.8477	&	128.2697	&	764.3297	&	269.5327	&	253.2762	&	222.7860	\\
Gaussian avg	&	0.0250	&	0.0217	&	521.3331	&	354.4588	&	1695.7987	&	666.5034	&	566.4133	&	567.0086	\\
pmed5	&	N/A	&	0.0004	&	1.4797	&	1.1839	&		&	3.8239	&	4.6917	&		4.290 \\
pmed20	&	N/A	&	0.0098	&	153.5358	&	121.9013	&	550.8604	&	240.8364	&	193.8426	&	189.6854	\\
pmed35	&	N/A	&	0.0666	&	3225.4469	&	1957.6519	&		&	1939.0170	&	2326.1185	&	1801.2031	\\
$p$-median avg	&	N/A	&	0.0243	&	1021.8706	&	652.1928	&		&	723.1452	&	857.3750	&	704.8966	\\ \hline
\end{tabular}
\label{fig:hier_results_comp}
\end{table*}
